Evidence of widspread maintenence of functional composition in vertebrate communities

### 2021-09-02 paper/manuscript.Rmd
Taking another stab at the discussion.

---
title: "Functional timeseries reveal no systemic shifts in vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Perry de Valpine"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |

@@

Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g. fish @trindade-santos2020, birds @jarzyna2016; @barnagaud2017].
<!-- so what do these studies say? does that go here? --> However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.

Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
We assess thousands of mammal, bird, and amphibian functional diversity timeseries to 1) determine how functional shifts compares to species change and 2) assess the types of functional shifts that are most common across communities.
<!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change. We do not therefore seek to classify the global trend in functional diversity change, but rather assess the utility of a functional lens in understanding biodiversity change.-->

@@

We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
There was also no temporal trend for studies broken down by taxa, biome or realm.
We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
Non-significant overall trends indicate that an equal number of studies experienced increasing and decreasing trends in those metrics.

@@

<!-- Our study represents the largest broad-scale multi-taxa assessment of functional change through time, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities. Consistent with previous species-based syntheses, we found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018]. While nearly all of the studies included in our data experienced a significant increase or decrease in species richness, trends were observed in equal numbers resulting in an overall trend not significantly different from zero. -->

<!-- Surprisingly, we did not detect an overall trend in any of the functional diversity metrics. Similar to species richness, communities are experiencing a wide range of functional shifts that balance each other out. While our dataset likely under-samples areas with most direct human impact, anthropogenic change is hypothesized to impact even the most remote systems. Conservatively, our findings therefore indicate that in unperturbed mammal, bird, and amphibian communities there is no evidence of systemic functional richness loss or functional change. A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset. -->

@@

Communities demonstrated almost universal maintenance of functional composition.
While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three experienced a significant shift in any functional dimension.

Individual studies in our dataset demonstrate the variety of species and functional shifts that can happen concurrently in a community.

With only one significant pairwise relationship between study-level functional divergence slopes and functional evenness slopes, all other possible combinations of species and functional change were observed in at least one study.
Notably, there is no relationship between species richness change and change in functional richness, meaning that communities experiencing species losses do not necessarily experience accompanying functional richness, and in some cases may even experience an increase.
<!-- need to put a disclaimer about what "increasing" means here - it's increased richness relative to the expectation for a community of that size --> These communities gain species that expand the boundaries of the trait space while losing interior species.

The overall trend of increasing turnover is consistent with work indicating higher than expected background rates of turnover in contemporary communities [@gotelli2017; @magurran2018].
However, many studies did not experience increasing turnover (65%), suggesting that it is not universal.
Even for those studies without a turnover trend, significant functional shifts occurred.
While some communities may be changing more rapidly than usual, that turnover is not necessarily an indication of specific kind of functional change.

@@

@@

Here's the prediction of the field, here is the data of the field, given that the data doesn't match the prediction it's on the field to resolve these issues.

Why do we find no significant trend in the global data despite extensive predictions to the contrary from both theory and local studies?
First, there are a variety of data limitations to consider:

This study represents best practices in biodiversity synthesis and functional diversity analysis, however methodological limitations may still be obscuring potential systematic shifts in functional composition.
First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
While both of these challenges are pervasive in Ecology generally and could have potentially insidious implications for synthesis work, we can only use the data we have.
We do however emphasize that our study included many timeseries that experienced a significant species loss (\~ 52%), indicating that this is not simply a synthesis of pristine systems.

@@

>>>>

---
title: "Evidence of widspread maintenence of functional composition in vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Perry de Valpine"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |

@@

Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
<!-- and computational power? --> The majority of work has therefore focused largely on system-specific studies with traits collected in situ.
Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g. fish @trindade-santos2020, birds @jarzyna2016; @barnagaud2017].
<!-- so what do these studies say? does that go here? --> However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.

Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
We assess thousands of mammal, bird, and amphibian functional diversity timeseries to 1) establish a general trend of functional change 2) ???.

@@

We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
There was also no temporal trend in those metrics for studies broken down by taxa, biome or realm.
We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
Non-significant overall trends indicate that an equal number of studies experienced increasing and decreasing trends in those metrics.

@@

@@

Communities demonstrated almost universal maintenance of functional composition.
While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three experienced a significant shift in any functional dimension.
This suggests certain characteristics of the functional space are conserved even in the face of significant change in species identity, namely the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
On average, species additions have similar functional characteristics as lost species and therefore maintain the functional space structure.
<!-- language to distinguish between "functional traits" and a species' role in the context of the functional space -->

-   Paragraph about analog communities?

-   Paragraph about implications for BEF work

Our work challenges assumptions that functional loss is the default state of all or even many communities.
While we do not directly assess the disturbance histories of included communities, many are from sites that experienced significant disturbance or manual treatment (see appendix for metadata of included studies).
<!-- check this, get numbers? --> This is good news for ecosystem function!
But we may also be measuring the wrong dimensions of the trait space or the wrong traits.
Can't really say BEF needs to measure function directly because these communities aren't losing species or functional diversity - turnover is either within regular bounds or if not (if more turnover than normal) it's still maintaining community structure.
And presuming that the original structure was good for function, maintaining it is also good for function!

**What are we going to say about the disturbance/environmental change backgrounds of these communities??**

While we did not directly assess the impact of potential disturbance or environmental change in the included communities

While there are clear impacts on functional makeup in scenarios of disturbance (what other scenarios do we see??)

-   some of these studies expereience a disturbance/treatment!

-   even for those that didn't it is remarkable that these metrics were conserved so universally!

    The overall trend of increasing turnover is consistent with work indicating higher than expected background rates of turnover in contemporary communities [@gotelli2017; @magurran2018].

@@

<!-- Here's the prediction of the field, here is the data of the field, given that the data doesn't match the prediction it's on the field to resolve these issues. -->

@@

Despite widespread concern about the loss of functional diversity, our analysis does not indicate significant change in functional composition is the rule for communities.
Why

This study represents the best available biodiversity and trait data paired with best practices in biodiversity synthesis.
Why then do we find no change despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? --> and and many local studies?
First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017] (other sampling citation!), and the subset of data in this study exhibits a strong Northern Hemisphere bias.
We may simply not have data from those areas experiencing the greatest functional shifts.
We do however emphasize that even disturbed sites can maintain structure (citation), and therefore the inclusion of these sites would not necessarily change the result.

@@

Notably, the metrics we calculated do not capture shifts in the location of the functional space.
For example two communities could have very similar metric values but no overlap in their trait spaces.
This is especially critical in the context of biodiversity change, as a species could be lost and replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar population attributes.
This may be particularly relevant for communties that exist in a rapidly changing envrionment (both abiotic and biotic), as communities may maintain some characteristcs of their functional composition while shifting to track the new conditions.
Metrics developed to assess shifts and overlaps in trait spaces are now being developed (citation), and further exploration of time series using those approaches may disentangle maintained characteristics from occupied trait space.
